---
title: "The Relationship Between Democracy Level and Voter Turnout"
author: "Hope Verciglio"
description: "This project aims to evaluate how voter turnout in a country is affected by its level of democracy."
output:
  distill::distill_article:
    self_contained: false
---



## Introduction
In this project, I will study different countries that are classified as electoral democracies and assess how these countries' democracy level scores impact their election voter turnout. 

### Hypothesis
My hypothesis is that citizens in countries classified as having a lower level of democracy are going to be less likely to vote than citizens who live in countries that have a higher democracy score. I predict that there will be a positive relationship between my variables. In other words, as the level of democracy increases then voter turnout will also increase. 

I predict this to be the relationship since lower levels of democracy are often linked to corruption which could jeopardize the legitimacy of elections. For citizens living in countries that hold elections that are not free and fair, they could perceive that their vote does not matter leading to lower levels of voter turnout. I believe that this topic is interesting because 100% voter turnout is incredibly rare, and I am interested in how there are some people who actively chose not to vote in their nation’s elections, despite the fact that the policies that their elected officials make will impact them. 


```{r}
library(tidyverse)
library(readxl)
library(readr)

Voter_Turnout <-  read_xlsx("data/idea_export_voter_turnout_database_world.xlsx")

Electoral_Democracy <- read_xlsx("data/List_of_Electoral_Democracies_FIW24.xlsx", skip = 1)

Freedom_House <- read_xlsx("data/All_data_FIW_2013-2025.xlsx", 
    sheet = "FIW13-25", skip = 1)
```

## Data
```{r}
voter_turnout_names <- Voter_Turnout |>
  select(-ISO2, -ISO3) |>
  rename(Country2 = Country) |>
  mutate(
    Country = case_when(
      `Country2`== "South Korea" ~ "Korea, Republic of",
      `Country2`=="Côte d'Ivoire" ~ "	Cote d'Ivoire",
      `Country2`=="Czechia" ~ "Czech Republic", 
      `Country2`=="Iran, Islamic Republic of" ~ "Iran", 
      `Country2`=="Korea, Republic of" ~ "South Korea",
      `Country2`=="Lao People's Dem. Republic" ~ "North Korea", 
      `Country2`=="North Macedonia, Republic of" ~ "North Macedonia", 
      `Country2`=="Republic of The Congo (Brazzaville)" ~ "	Congo (Brazzaville)", 
      `Country2`=="Russian Federation" ~ "Russia",
      `Country2`=="São Tomé and Príncipe" ~ "Sao Tome and Principe", 
      `Country2`=="Syrian Arab Republic" ~ "Syria", 
      `Country2`=="Tanzania, United Republic of" ~ "Tanzania",
      `Country2`=="Viet Nam" ~ "Vietnam",
      .default = `Country2`))

Electoral_Democracy_new <- Electoral_Democracy |>
  rename(`Electoral Democracy` = `Electoral Democracy Designation in FIW 2025`)

Joined_data_clean <- Electoral_Democracy_new |>
  inner_join(voter_turnout_names, by = "Country") |>
  select(-Country2) |>
  mutate(`Voter Turnout` = as.numeric(str_remove(`Voter Turnout`, "%")))

Freedom_House <- Freedom_House |>
  rename(Country = `Country/Territory`)

Joined_data_extra_clean <- Joined_data_clean |>
  left_join(Freedom_House |> select(Country, Total), by = "Country") |>
  rename(`Democracy Score` = Total)
```
### Data Sources
My data set called “Voter Turnout” comes from the International Institute for Democracy and Electoral Assistance (International IDEA). It includes the voter turnout(as a percent) as well as the dates for each election for many countries. 

My data set called “Electoral Democracy” also comes from Freedom House, and it states which of the 195 countries in the world are classified as electoral democracies in 2025. Since this project is studying voter turnout, I used this data to filter out the countries that do not hold elections as they would not provide any data related to voter turnout. 

My data set called “Freedom House” comes from Freedom House. Freedom House is a U.S.-based, nonpartisan non-governmental organization that advocates for democracy, political freedom, and human rights worldwide. It annually assesses the level of political rights and civil liberties in the nations and territories of the world. This data features several different scores that countries received from Freedom House’s “Freedom in the World” report based on items like electoral process, freedom of expression, personal autonomy, and individual rights. This data was used to help me identify from the countries that hold elections which ones lean more democratic. For the point of this project I looked at the total score for each country. This variable is the aggregate score for all categories accessing democratic level. 

### Dependent and Independent Variables 
My independent variable is the total aggregate democratic score level from the “Freedom House” data set. This variable is on a scale from 0 to 100, where 100 represents the highest level of freedom and 0 the lowest. 

My dependent variable is the voter turnout in a country from my data set “Voter Turnout”. This variable is measured as a percentage.  

### Criteria For My Hypothesis 
Since my hypothesis is that as the level of democracy increases then voter turnout will also increase, a pattern that would support my hypothesis would be lower voter turnout in countries with lower total democratic level scores and higher voter turnout in countries with greater democratic scores. 


```{r}
ggplot(Joined_data_clean, aes(x = `Electoral Democracy`, y = `Voter Turnout`, fill = `Electoral Democracy`)) +
  geom_boxplot() +
  labs(
    title = "Voter Turnout by Electoral Democracy Status",
    x = "Electoral Democracy Designation (FIW 2025)",
    y = "Voter Turnout (%)"
  )
```
```{r}
Non_Democracy_Mean_Turnout <- Joined_data_clean |>
  filter(`Electoral Democracy`== "No") |>
  summarize(non_democracy_turnout = mean(`Voter Turnout`, na.rm = TRUE)) 

Democracy_Mean_Turnout <- Joined_data_clean |>
  filter(`Electoral Democracy`== "Yes") |>
  summarize(democracy_turnout = mean(`Voter Turnout`, na.rm = TRUE)) 


Mean_Turnouts <- bind_cols(Non_Democracy_Mean_Turnout, Democracy_Mean_Turnout)
Mean_Turnouts

knitr::kable(Mean_Turnouts, digits = 2, col.names = c("Non Democracy Mean Turnout", "Democracy Mean Turnout"))
```
## Table and Graph Anaylsis 
Based on the boxplot for voter turnout by electoral democracy status and the table of mean voter turnout for democracies and non-democracies, countries that are classified as democracies have higher levels of voter participation. In the boxplot we can see that the median voter turnout for electoral democracies is higher than the median of non-electoral democracies voter turnout. This is also represented in the table of mean voter turnout, since electoral democracies had a greater mean value of voter turnout than non-electoral democracies did. 


```{r}
regression_plot <- ggplot(Joined_data_extra_clean, aes(x = `Democracy Score`, y = `Voter Turnout`)) +
geom_point(color = "steelblue1") +
  geom_smooth(method = "lm", se = FALSE, color = "indianred1", size = 1.5)
regression_plot
```
```{r}
regression <- lm(`Voter Turnout` ~ `Democracy Score`, data = Joined_data_extra_clean)
regression

summary(regression)
```
## Regression Analysis 
Based on this regression line, we can see that there is a positive relationship between a country's democracy score and its voter turnout. The y-intercept is 59.49, which in this context says that the baseline of voter turnout is around 59.49 percent. In addition, for every one point increase in a country's democracy score given to it by Freedom House(where 1 is a low democracy score and 100 is a high democracy score) voter turnout is expected to increase by 0.1634. While this is a positive relationship, the increase in voter turnout based on the democracy score appears relatively small. 
